% make clean all
% make site deploy
%\documentclass[12pt,titlepage]{article}
\documentclass{tufte-book}
\usepackage{url}
% \usepackage[authoryear]{natbib}

\usepackage{multicol} %%For eqn sheet https://tex.stackexchange.com/questions/152093/troubles-with-a-two-column-document-using-tufte-handout

\usepackage{graphicx}
\usepackage{animate}  %https://help.geogebra.org/topic/how-to-include-animated-gif-in-latex-pdf

\bibliographystyle{genetics}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{wasysym}
\usepackage{array}
%\usepackage[normalem]{ulem}   %%MArs/Venus symbols
%\usepackage{marginnote}
\usepackage{color,booktabs}
%\usepackage{fullpage}
\usepackage{nicefrac}
\usepackage{csvsimple}
%\usepackage[hypertexnames=false]{hyperref}
\hypersetup{  colorlinks=false,% hyperlinks will be black
  linkbordercolor=red,% hyperlink borders will be red
  citebordercolor=blue,
  pdfborderstyle={/S/U/W 1}, % border style will be underline of width 1pt
  hypertexnames=false   %TO fix issue of TOC renew chapter issue https://tex.stackexchange.com/questions/231361/table-of-contents-labels-not-directing-hyperlinks-to-correct-page
}  %{colorlinks=true, urlcolor=black, citecolor=black, linkcolor=black,linkbordercolor=black}
%\usepackage[normalem]{ulem}

%\usepackage{lineno}  %%line numbering

\usepackage{xeCJK}
%\setCJKmainfont{MingLiU}
%\setCJKsansfont{Microsoft JhengHei}
%% chinese characters https://tex.stackexchange.com/questions/200449/how-to-input-traditional-chinese-in-pdflatex

%\usepackage{slashbox}  %for diagonal slash table https://tex.stackexchange.com/questions/27193/latex-table-cell-with-a-diagonal-line-and-2-sub-cells

\newcommand{\ec}[1]{{\em \color{blue} #1}}

\newcommand{\gc}[1]{{\color{green} #1}}  %\color{black} #1}}   %
\newcommand{\sa}[1]{{\color{black} #1}}
\newcommand{\plr}[1]{{\em \color{green} (#1)}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\half}{\tfrac{1}{2}}

%FOR compiling on xelatex
\renewcommand\allcapsspacing[1]{{\addfontfeature{LetterSpace=15}#1}}
%%comment out this line for overleaf
\renewcommand\smallcapsspacing[1]{{\addfontfeature{LetterSpace=10}#1}} %%comment out this line for overleaf
\usepackage[colorinlistoftodos  ,disable]{todonotes}  % ,disable  ,disable


%\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{multirow}

\newcommand{\wbar}{\overline{w}}
% New commands added by Simon:
\newcommand{\eqn}{eqn}
\newcommand{\fis}{F_{\mathrm{IS}}}
\newcommand{\fit}{F_{\mathrm{IT}}}
\newcommand{\fst}{F_{\mathrm{ST}}}
\newcommand{\Wbar}{\overline{W}}
\newcommand{\dNdS}{\nicefrac{d_N}{d_S}}

%\newcommand{\graham}[1]{\todo[size=\scriptsize, color=red!50]{#1}}
%Image from the Biodiversity Heritage Library. Contributed by [Holding Institution]. | www.biodiversitylibrary.org
%Image from the Biodiversity Heritage Library. Contributed by [Holding
%Institution]. licensed under CC BY 2.0

\usepackage{caption}
\newcommand{\BHLCC}[4]{\newline \noindent \tiny{#1 Image from the \href{#2}{Biodiversity
      Heritage Library}. Contributed by #3. Licensed under CC
    BY-#4. \par}} %\caption*{
\newcommand{\BHLNKC}[3]{\newline \noindent  \tiny{#1 Image from the \href{#2}{Biodiversity
      Heritage Library}. Contributed by #3. No known copyright restrictions.}}
\newcommand{\BHLNC}[3]{\newline \noindent \tiny{#1 Image from the \href{#2}{Biodiversity Heritage Library}. Contributed by #3. Not in copyright.}}
\newcommand{\IANC}[3]{\newline \noindent \tiny{#1 Image from the \href{#2}{Internet Archive}. Contributed by #3. No known copyright restrictions.}}

\newcommand{\wikimedia}[4]{\newline \noindent \tiny{#1 Image from
    \href{#2}{wikimedia} contributed by #3. #4 }}  %%use #4 slot for
                                %%public domain etc


\newcommand{\gitcode}[1]{Code \href{#1}{here}.}
\newcommand{\PLOSccBY}{{licensed under CC BY 4.0}}


\newcommand{\graham}[1]{\todo[size=\scriptsize, color=red!50]{#1}}

\newcommand{\erin}[1]{\todo[size=\scriptsize, color=blue!50]{#1}}

%\newcounter{Question}[section]
%\setcounter{Question}{1}

\newcounter{question}[chapter]  %[section]   %%modified from https://www.sharelatex.com/learn/Counters

%https://tex.stackexchange.com/questions/428652/newenvironment-for-tcolorbox-and-lstlisting
\usepackage[most]{tcolorbox} %\usepackage{tcolorbox}

%mybox
\newtcolorbox[use counter=question]{question}[1]{colback=red!5!white,
colframe=red!75!black,fonttitle=\bfseries, 
title= Question~\thequestion. #1}

\newenvironment{ChapterSummary}{%
\begin{center} {\large {\bf Chapter
      \thechapter~summary.}} \end{center}
%We described:
\begin{itemize}}{\end{itemize}}
\tcolorboxenvironment{ChapterSummary}{ colback=blue!5!white,
colframe=blue!75!black
}

%\newtcbox{\mybox}[1]{colback=red!5!white,
%colframe=red!75!black,fonttitle=\bfseries,
%title= Question~\thequestion. #1}

%\newenvironment{question}[1][] {\refstepcounter{question}\par\medskip
 % \textbf{Question~\thequestion. #1} \rmfamily}{\medskip}

%\newenvironment{question}[1][] {\refstepcounter{question}\par   \begin{tcolorbox}
 %   \medskip
  %  \textbf{Question~\thequestion. #1}\rmfamily \end{tcolorbox} 
%}{ \medskip}  %\end{tcolorbox} 

%\tcolorboxenvironment{question}{colback=red!5!white,
%colframe=red!75!black}

\setcounter{secnumdepth}{2}


%\newenvironment{question}[1][]{\refstepcounter{question}\par\medskip
%   \textbf{Question~\thequestion. #1} \rmfamily}{\medskip}


% Prints the month name (e.g., January) and the year (e.g., 2008)
\newcommand{\monthyear}{%
  \ifcase\month\or January\or February\or March\or April\or May\or June\or
  July\or August\or September\or October\or November\or
  December\fi\space\number\year
}


\definecolor{rev1}{rgb}{1, 0, 0}


\begin{document}

\title{Population and\newline Quantitative \newline Genetics}
\author{Graham Coop}
%  \small $^1$ Department of Evolution and Ecology \& Center for Population Biology,\\
%  \small University of California, Davis.\\
%  \small To whom correspondence should be addressed: \texttt{gmcoop@ucdavis.edu}\\
%  \small This work is licensed under a Creative Commons Attribution 3.0 Unported License.\\
%  \small http://creativecommons.org/licenses/by/3.0/ \\
%  \small i.e. you are free to reuse and remix this work, but please include an attribution to the original.


\date{}
\maketitle
\newpage

\begin{fullwidth}
~\vfill
\thispagestyle{empty}
\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}
%Copyright \copyright\ \the\year\
Author: \thanklessauthor\\
  \small Author address: Department of Evolution and Ecology \& Center for Population Biology,\\
 \small University of California, Davis.\\
  \small To whom correspondence should be addressed: \texttt{gmcoop@ucdavis.edu}\\
 \small This work is licensed under a Creative Commons Attribution 3.0 Unported License.\\
 \small http://creativecommons.org/licenses/by/3.0/ \\
 \small i.e. you are free to reuse and remix this work, but please include an attribution to the original.
\par  Typeset using \LaTeX~and the \smallcaps{tufte-latex} book style.
\par  The \LaTeX~code and R code for this book are kept here
https://github.com/cooplab/popgen-notes/ and again are under a Creative Commons Attribution 3.0 Unported License.
\par\textit{Updated on \monthyear}
\end{fullwidth}

\newpage

\begin{fullwidth}
~\vfill
\thispagestyle{empty}

% r.7 dedication
%\cleardoublepage
~\vfill
\begin{doublespace}
\noindent\fontsize{16}{18}\selectfont\itshape   %{18}{22}\selectfont\itshape
\nohyphenation
\begin{center}
Dedicated to my parents, partner, and children. Thank you for teaching me the
true joy of descent with modification.
\end{center}
\end{doublespace}
\vfill
\vfill

\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}
%Copyright \copyright\ \the\year\
%Author: \thanklessauthor\\
  \small This text and the latex code are released under the
  Creative Commons Attribution 3.0 Unported License. This book was developed from my set of notes for the \href{http://www-eve.ucdavis.edu/eve/pbg/}{Population
  Biology graduate group} core class (PBG200A) and Undergraduate Population and
  Quantitative Genetics class (EVE102) at UC Davis. Thanks to the many
  students who've read these notes and suggested improvements. Thanks
 to Simon Aeschbacher, Vince Buffalo, Erin Calfee, and Michael Whitlock who read
 and extensively edited earlier drafts of these notes. To illustrate these notes, I've used old scientific and natural history illustrations, in part
 because they are out of copyright but mainly because they bring me
 joy. Many of the old images come from
 \href{https://www.biodiversitylibrary.org/}{Biodiversity Heritage
   Library}--- a consortium of natural history institutions that are
 digitizing their collections and make them freely available
 online. If you enjoy the images consider donating to the
 \href{http://library.si.edu/donate-bhl}{BHL}. Many of the data and
 simulation graphics in the book were prepared in \citet{Rstats},
 the code for each is linked to from the caption of each figure. In many cases
 data were extracted from old figures using the
 \href{https://automeris.io/WebPlotDigitizer/}{WebPlotDigitizer}
 tool, as such I advise re-extracting the data if you wish to use it
 for research purposes.
\end{fullwidth}

%\newpage
\tableofcontents

%\linenumbers
%\modulolinenumbers[2]
\include{math_background/Equation_sheet}



%\newpage
\include{Chapters/intro}
\include{Chapters/eugenics_frontpiece}
\include{Chapters/chapter-01}
\include{Chapters/Population_structure}
\include{Chapters/genetic_drift}
\include{Chapters/Molecular_evolution}
\include{Chapters/genetic_drift_pop_structure}
\include{Chapters/Pheno_var}
\include{Chapters/Response_to_sel}
\include{Chapters/Multi_trait_selection}
\include{Chapters/One_locus_selection}
\include{Chapters/Interaction_selection_mut_mig}
\include{Chapters/Genetic_drift_selection}
%\include{chapter-07}
\include{Chapters/Linked_selection}
\include{Chapters/multi_locus_sel}

\setcounter{figure}{0}
\setcounter{section}{0}
\setcounter{equation}{0}
\setcounter{chapter}{0}
\renewcommand{\thechapter}{\Alph{chapter}.}
\renewcommand{\theequation}{\Alph{chapter}.\arabic{equation}}
\renewcommand{\thesection}{\Alph{chapter}.\arabic{section}}
\renewcommand{\thefigure}{\Alph{chapter}.\arabic{figure}}

\include{math_background/Math-background}

%\renewcommand{\thechapter}{}  %{\Alph{chapter}.}
\bibliography{popgen_notes}

\end{document}
